Aligning species range data to better serve science and conservation

### 2015-12-21 data_explore/data_explore.Rmd
trying this again... cleaning up data_explore plots and .csvs

library(ggplot2)
library(maps)
library(readr)
library(data.table)
library(tidyr)
library(dplyr)
library(stringr)

@@

dir_anx <- file.path(dir_N, 'git-annex/globalprep/SPP_ICO')
dir_fig <- file.path(getwd(), 'figures')

source('R/data_explore_fxns.R')
```

@@

### LOICZID cell ID numbers - used in plotting rasters
loiczid_raster_file  <- 'shiny/data/loiczid_raster.grd'
loiczid_raster       <- raster(loiczid_raster_file)
names(loiczid_raster) <- 'loiczid'
```

```{r get_spp_list, echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE}
data_file <- 'data/spp_list_w_area_percent_noextinct_2015.csv'

if (!file.exists(data_file)) {
  ### species info list - scinames, IUCN species ID, AM species ID, category etc
  spp_list_base <- read.csv('data/spp_am_v_iucn_2015_inclDD.csv', stringsAsFactors = FALSE)

@@

spp_list <- read.csv(data_file, stringsAsFactors = FALSE) %>%
  filter(!is.na(sm_perc)) 
  ### will ditch the ones where no IUCN polygon exists - or are these species
  ### whose ranges only occur in cells beyond the LOICZID grid?
spp_list <- spp_list %>%
  left_join(read.csv('sciname_lookups/spp_groupnames.csv', stringsAsFactors = FALSE),
            by = 'spp_group')

@@

``` {r child = 'R/data_explore_pt1.Rmd', eval = FALSE}
```
<!-- # bar chart: number of spp per group
# scatter plot: area compare IUCN vs AM, including log plot and plot by species group
# rangemap: Chromis alpha -->

``` {r, child = 'R/data_explore_pt2.Rmd', eval = FALSE}
```

@@

``` {r child = 'R/data_explore_pt3.Rmd', eval = FALSE}
```
<!-- # Multiple species maps per quadrant -->

@@

data_file <- 'data/spp_list_w_area_percent_noextinct_2015.csv'
spp_map_pairs <- read.csv(data_file, stringsAsFactors = FALSE) %>%
  filter(!is.na(popn_category)) %>% ### filter out unassessed species
  filter(!is.na(sm_perc))           ### filter out NA maps (usually IUCN maps with no polygons)

@@

spp_all_file <- 'data/spp_all_2015.csv'
spp_all <- read.csv(spp_all_file, stringsAsFactors = FALSE)

iucn_marine_file <- 'data/spp_iucn_marine.csv'
spp_iucn_marine <- read.csv(iucn_marine_file, stringsAsFactors = FALSE)

iucn_maps_file <- 'data/spp_iucn_maps.csv'
spp_iucn_maps <- read.csv(iucn_maps_file, stringsAsFactors = FALSE) %>%
  filter(spp_group != 'hagfishes') %>%           ### not from IUCN...
  filter(sciname %in% spp_iucn_marine$sciname)   ### ditch freshwater spp

spp_am_file <- 'data/spp_am.csv'
spp_am <- read.csv(spp_am_file, stringsAsFactors = FALSE)

@@

sciname_lookup <- read.csv('sciname_lookups/spp_iucn_all.csv', stringsAsFactors = FALSE)

@@

taxa_df_long2 <- taxa_df %>%
  dplyr::select(class, AquaMaps = n_am_class_solo, IUCN = n_iucn_class_solo, Both = n_class_both) %>%
  gather(dataset, n_spp, AquaMaps:Both) %>%
  group_by(class) %>%
  mutate(n_total = sum(n_spp, na.rm = TRUE),
         percent = n_spp/n_total * 100)
bar_taxa2 <- ggplot(taxa_df_long2, aes(x = class, y = percent, fill = dataset)) +
  ggtheme_plot + 
  geom_bar(stat = 'identity', alpha = .5) +
  scale_fill_manual(values = c('AquaMaps' = 'red', 'Both' = 'mediumorchid', 'IUCN' = 'blue')) +
  geom_text(aes(label = sprintf('n = %s', n_total), y = 3), 
            size = 3, angle = 90, hjust = 0, color = 'grey30') +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)) + 
  scale_y_continuous(expand = c(0, 0), limits = c(0, 100))
  labs(x = 'Class', y = 'Percent of mapped species',
       title = 'Distribution of species maps by taxonomic group (class)')
print(bar_taxa2)
ggsave(filename = file.path(dir_fig, 'bar_by_taxa_all.png'), height = 4, width = 8)

taxa_df_long3 <- taxa_df %>%
  dplyr::select(class, AquaMaps = n_am_class_solo_ass, IUCN = n_iucn_class_solo, Both = n_class_both) %>%
  gather(dataset, n_spp, AquaMaps:Both) %>%
  group_by(class) %>%
  mutate(n_total = sum(n_spp, na.rm = TRUE),
         percent = n_spp/n_total * 100)
bar_taxa3 <- ggplot(taxa_df_long3, aes(x = class, y = percent, fill = dataset)) +
  ggtheme_plot + 
  geom_bar(stat = 'identity', alpha = .5) +
  scale_fill_manual(values = c('AquaMaps' = 'red', 'Both' = 'mediumorchid', 'IUCN' = 'blue')) +
  geom_text(aes(label = sprintf('n = %s', n_total), y = 3), 
            size = 3, angle = 90, hjust = 0, color = 'grey30') +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)) + 
  scale_y_continuous(expand = c(0, 0), limits = c(0, 100))
  labs(x = 'Class', y = 'Percent of mapped and assessed species',
       title = 'Distribution of species maps by taxonomic group (class)')
print(bar_taxa3)
ggsave(filename = file.path(dir_fig, 'bar_by_taxa_assessed.png'), height = 4, width = 8)

@@

``` {r, child = 'R/data_explore_global_maps.Rmd', eval = FALSE}
```

``` {r, child = 'R/data_explore_am_thresh.Rmd', eval = FALSE}
```

## Changes in SPP for scenarios

Comparing three scenarios to the current (v2015) global model for SPP:

* Baseline is: preferring IUCN rangemaps over Aquamaps rangemaps; Aquamaps 'presence' threshold is set at 40% or greater.
* Scenario 1: preferring IUCN rangemaps over Aquamaps; AM presence threshold set at >= 1% (effectively, all occurrences).
* Scenario 2: preferring AM maps over IUCN; AM presence >= 40%.
* Scenario 3: preferring AM over IUCN; AM presence >= 1%.

Boxplots:

![Absolute change in SPP score](figures/SPP_scenarios/boxplot_abs_diff.png)

![Relative change in SPP score](figures/SPP_scenarios/boxplot_rel_diff.png)

>>>>

library(ggplot2)
library(maps)
library(readr)
library(data.table)
library(raster)
library(tidyr)
library(dplyr)
library(stringr)

@@

dir_git <- '~/github/IUCN-AquaMaps'

dir_anx <- file.path(dir_N, 'git-annex/globalprep/SPP_ICO')
dir_fig <- file.path(dir_git, 'figures')
dir_data <- file.path(dir_git, 'data')

source('data_explore_fxns.R')
### When knitting this, it automatically sets WD to be this directory...
```

@@

### LOICZID cell ID numbers - used in plotting rasters
loiczid_raster_file  <- file.path(dir_git, 'shiny/data/loiczid_raster.grd')
loiczid_raster       <- raster(loiczid_raster_file)
names(loiczid_raster) <- 'loiczid'
```

```{r get_spp_list, echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE}
data_file <- file.path(dir_data, 'spp_list_w_area_all.csv')

if (!file.exists(data_file)) {
  ### species info list - scinames, IUCN species ID, AM species ID, category etc
  spp_list_base <- read.csv(file.path(dir_data, 'spp_am_v_iucn.csv'),
                            stringsAsFactors = FALSE)

@@

spp_list <- read.csv(data_file, stringsAsFactors = FALSE) 
### At this point, spp_list contains subpops and aliases, and zero-area IUCNs

spp_list <- spp_list %>%
  filter(!is.na(sm_perc)) %>%
  ### will ditch the ones where no IUCN polygon exists - or are these species
  ### whose ranges only occur in cells beyond the LOICZID grid?
  filter(!str_detect(spatial_source, 'subpop') & !str_detect(spatial_source, 'alias'))
spp_list <- spp_list %>%
  left_join(read.csv(file.path(dir_data, 'sciname_lookups/spp_groupnames.csv'), 
                     stringsAsFactors = FALSE),
            by = 'spp_group')

write.csv(spp_list, file.path(dir_data, 'spp_list_w_area_trimmed.csv'))

@@

``` {r child = 'data_explore_pt1.Rmd', eval = FALSE}
```
<!-- # bar chart: number of spp per group
# scatter plot: area compare IUCN vs AM, including log plot and plot by species group
# rangemap: Chromis alpha -->

``` {r, child = 'data_explore_pt2.Rmd', eval = FALSE}
```

@@

<!-- # Multiple species maps per quadrant -->
``` {r child = 'data_explore_pt3.Rmd', eval = FALSE}
```

@@

data_file <- file.path(dir_data, 'spp_list_w_area_trimmed.csv')
spp_map_pairs <- read.csv(data_file, stringsAsFactors = FALSE) %>%
  filter(!is.na(popn_category)) %>% ### filter out unassessed species
  filter(!is.na(sm_perc))           ### filter out NA maps (usually IUCN maps with no polygons)

@@

spp_all_file <- file.path(dir_data, 'spp_all.csv')
spp_all <- read.csv(spp_all_file, stringsAsFactors = FALSE)

iucn_marine_file <- file.path(dir_data, 'spp_iucn_marine.csv')
spp_iucn_marine <- read.csv(iucn_marine_file, stringsAsFactors = FALSE)

message('Getting IUCN maps list')
iucn_maps_file <- file.path(dir_data, 'spp_iucn_maps.csv')
spp_iucn_maps <- read.csv(iucn_maps_file, stringsAsFactors = FALSE) %>%
  filter(spp_group != 'hagfishes') %>%           ### not from IUCN...
  filter(sciname %in% spp_iucn_marine$sciname)   ### ditch freshwater spp

spp_am_file <- file.path(dir_data, 'spp_am.csv')
spp_am <- read.csv(spp_am_file, stringsAsFactors = FALSE)

@@

sciname_lookup <- read.csv(file.path(dir_data, 'sciname_lookups/spp_iucn_scinames.csv'), stringsAsFactors = FALSE)

@@

taxa_df_long2 <- taxa_df %>%
  dplyr::select(class, AquaMaps = n_am_class_solo, IUCN = n_iucn_class_solo, Both = n_class_both) %>%
  gather(dataset, n_spp, AquaMaps:Both) %>%
  group_by(class) %>%
  mutate(n_total = sum(n_spp, na.rm = TRUE),
         percent = n_spp/n_total * 100)
bar_taxa2 <- ggplot(taxa_df_long2, aes(x = class, y = percent, fill = dataset)) +
  ggtheme_plot + 
  geom_bar(stat = 'identity', alpha = .5) +
  scale_fill_manual(values = c('AquaMaps' = 'red', 'Both' = 'mediumorchid', 'IUCN' = 'blue')) +
  geom_text(aes(label = sprintf('n = %s', n_total), y = 3), 
            size = 3, angle = 90, hjust = 0, color = 'grey30') +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)) + 
  scale_y_continuous(expand = c(0, 0), limits = c(0, 100)) + 
  labs(x = 'Class', y = 'Percent of mapped species',
       title = 'Distribution of species maps by taxonomic group (class)')
print(bar_taxa2)
ggsave(filename = file.path(dir_fig, 'bar_by_taxa_all.png'), height = 4, width = 8)

taxa_df_long3 <- taxa_df %>%
  dplyr::select(class, AquaMaps = n_am_class_solo_ass, IUCN = n_iucn_class_solo, Both = n_class_both) %>%
  gather(dataset, n_spp, AquaMaps:Both) %>%
  group_by(class) %>%
  mutate(n_total = sum(n_spp, na.rm = TRUE),
         percent = n_spp/n_total * 100)
bar_taxa3 <- ggplot(taxa_df_long3, aes(x = class, y = percent, fill = dataset)) +
  ggtheme_plot + 
  geom_bar(stat = 'identity', alpha = .5) +
  scale_fill_manual(values = c('AquaMaps' = 'red', 'Both' = 'mediumorchid', 'IUCN' = 'blue')) +
  geom_text(aes(label = sprintf('n = %s', n_total), y = 3), 
            size = 3, angle = 90, hjust = 0, color = 'grey30') +
  theme(axis.text.x = element_text(angle = 75, hjust = 1)) + 
  scale_y_continuous(expand = c(0, 0), limits = c(0, 100)) +
  labs(x = 'Class', y = 'Percent of mapped and assessed species',
       title = 'Distribution of species maps by taxonomic group (class)')
print(bar_taxa3)
ggsave(filename = file.path(dir_fig, 'bar_by_taxa_assessed.png'), height = 4, width = 8)

@@

``` {r, child = 'data_explore_global_maps.Rmd', echo = FALSE, message = FALSE, eval = TRUE}
```

``` {r, child = 'data_explore_am_thresh.Rmd', echo = FALSE, message = FALSE, eval = TRUE}
```

### 2015-12-21 data_explore/data_explore_global_maps.Rmd
trying this again... cleaning up data_explore plots and .csvs

``` {r setup, message = FALSE, eval = FALSE}
library(rasterVis)
library(ggplot2)
library(maps)
library(raster)
library(data.table)
library(tidyr)
library(dplyr)
library(stringr)
library(readr)

dir_N <- c('Windows' = '//neptune.nceas.ucsb.edu/data_edit',
           'Darwin'  = '/Volumes/data_edit',
           'Linux'   = '/var/data/ohi')[[ Sys.info()[['sysname']] ]]

dir_anx <- file.path(dir_N, 'git-annex/globalprep/SPP_ICO')
dir_am_data <- file.path(dir_N, 'git-annex/globalprep/_raw_data/aquamaps/v2015/csv')
dir_fig <- file.path(getwd(), 'figures')

source('R/data_explore_fxns.R')

data_file <- 'data/spp_list_w_area_percent_noextinct_2015.csv'

spp_list <- read.csv(data_file, stringsAsFactors = FALSE) %>%
  filter(!is.na(sm_perc)) ### will ditch the ones where no IUCN polygon exists

``` {r load data sets, echo = FALSE, message = FALSE}
iucn_spp_cells <- read_csv(file.path(dir_anx, 'explore_am_v_iucn/iucn_spp_cells_2015.csv'), 
                           col_types = 'cdddd', progress = TRUE) %>%
  as.data.frame() %>%
  filter(presence != 5) ### removes extinct subpopulations

am_spp_cells   <- read_csv(file.path(dir_anx, 'explore_am_v_iucn/am_spp_cells_2015.csv'), progress = TRUE)

### LOICZID cell ID numbers - used in plotting rasters
loiczid_raster_file  <- 'shiny/data/loiczid_raster.grd'
loiczid_raster       <- raster(loiczid_raster_file)
names(loiczid_raster) <- 'loiczid'
```

```{r global map matched species, echo = FALSE, warning = FALSE, message = FALSE}

@@

### Number of aquamaps species per cell
iucn_pair_rast_file <- 'rasters/iucn_pair_rast.tif'
am_pair_rast_file   <- 'rasters/am_pair_rast.tif'

@@

all_am_rast_file   <- 'rasters/am_all_rast.tif'

@@

levelplot(all_am_map, par.settings = myTheme, main = 'Entire Aquamaps Dataset')

all_am_ln_map <- log(all_am_map + 1)
all_am_ln_map <- all_am_ln_map/maxValue(all_am_ln_map)
levelplot(all_am_ln_map, par.settings = myTheme, main = 'Entire Aquamaps Dataset: log-transformed relative species richness', 
          colorkey = list(at = seq(0, 1, length = 15)))

am_ln_pts <- as.data.frame(rasterToPoints(all_am_ln_map))
spp_richness_am_plot <- ggplot(am_ln_pts, aes(x = x, y = y)) +
  ggtheme_map + 
  geom_raster(aes(fill = layer), alpha = .7) +
  scale_fill_continuous(low = 'white', high = 'blue') +
  borders('world', color='gray40', fill='gray45', size = .1) +  # create a layer of borders
  scale_x_continuous(breaks = seq(-180, 180, by = 30), expand = c(0, 2)) +
  scale_y_continuous(breaks = seq( -90,  90, by = 30), expand = c(0, 2)) +
  labs(title = 'AquaMaps Dataset: log-transformed relative species richness', x = NULL, y = NULL, fill = NULL)

spp_richness_am_plot
ggsave(file.path(dir_fig, 'all_am_ln_map.png'), height = 4, width = 8)

@@

all_iucn_rast_file   <- 'rasters/iucn_all_rast.tif'

@@

levelplot(all_iucn_map, par.settings = myTheme, main = 'Entire IUCN Dataset')

all_iucn_ln_map <- log(all_iucn_map + 1)
all_iucn_ln_map <- all_iucn_ln_map/maxValue(all_iucn_ln_map)
levelplot(all_iucn_ln_map, 
          par.settings = myTheme, 
          margin = list(axis = TRUE),
          main = 'Entire IUCN Dataset: log-transformed relative species richness', 
          colorkey = list(at = seq(0, 1, length = 15)))

iucn_ln_pts <- as.data.frame(rasterToPoints(all_iucn_ln_map))
spp_richness_iucn_plot <- ggplot(iucn_ln_pts, aes(x = x, y = y)) +
  ggtheme_map + 
  geom_raster(aes(fill = layer), alpha = .7) +
  scale_fill_continuous(low = 'white', high = 'blue') +
  borders('world', color='gray40', fill='gray45', size = .1) +  # create a layer of borders
  scale_x_continuous(breaks = seq(-180, 180, by = 30), expand = c(0, 2)) +
  scale_y_continuous(breaks = seq( -90,  90, by = 30), expand = c(0, 2)) +
  labs(title = 'IUCN Dataset: log-transformed relative species richness', x = NULL, y = NULL, fill = NULL)

spp_richness_iucn_plot
ggsave(file.path(dir_fig, 'all_iucn_ln_map.png'), height = 4, width = 8)

@@

![image](figures/all_aquamaps_map.png)

>>>>

```{r global map matched species, echo = FALSE, warning = FALSE, message = FALSE, eval = FALSE}

@@

### Number of aquamaps species per cell
iucn_pair_rast_file <- file.path(dir_data, 'rasters/iucn_pair_rast.tif')
am_pair_rast_file   <- file.path(dir_data, 'rasters/am_pair_rast.tif')

@@

all_am_rast_file   <- file.path(dir_data, 'rasters/am_all_rast.tif')

@@

am_levelplot1 <- levelplot(all_am_map, par.settings = myTheme, main = 'Entire Aquamaps Dataset')

all_am_ln_map <- log(all_am_map + 1)
all_am_ln_map <- all_am_ln_map/maxValue(all_am_ln_map)
am_levelplot1 <- levelplot(all_am_ln_map, par.settings = myTheme, main = 'Entire Aquamaps Dataset: log-transformed relative species richness', 
          colorkey = list(at = seq(0, 1, length = 15)))

print(am_levelplot1)
print(am_levelplot2)

# am_ln_pts <- as.data.frame(rasterToPoints(all_am_ln_map))
# spp_richness_am_plot <- ggplot(am_ln_pts, aes(x = x, y = y)) +
#   ggtheme_map + 
#   geom_raster(aes(fill = layer), alpha = .7) +
#   scale_fill_continuous(low = 'white', high = 'blue') +
#   borders('world', color='gray40', fill='gray45', size = .1) +  # create a layer of borders
#   scale_x_continuous(breaks = seq(-180, 180, by = 30), expand = c(0, 2)) +
#   scale_y_continuous(breaks = seq( -90,  90, by = 30), expand = c(0, 2)) +
#   labs(title = 'AquaMaps Dataset: log-transformed relative species richness', x = NULL, y = NULL, fill = NULL) 
# 
# spp_richness_am_plot
# ggsave(file.path(dir_fig, 'all_am_ln_map.png'), height = 4, width = 8)

@@

```{r all_iucn, eval = FALSE, echo = FALSE}

all_iucn_rast_file   <- file.path(dir_data, 'rasters/iucn_all_rast.tif')

@@

iucn_levelplot1 <- levelplot(all_iucn_map, par.settings = myTheme, main = 'Entire IUCN Dataset')

all_iucn_ln_map <- log(all_iucn_map + 1)
all_iucn_ln_map <- all_iucn_ln_map/maxValue(all_iucn_ln_map)
iucn_levelplot2 <- levelplot(all_iucn_ln_map, 
          par.settings = myTheme, 
          margin = list(axis = TRUE),
          main = 'Entire IUCN Dataset: log-transformed relative species richness', 
          colorkey = list(at = seq(0, 1, length = 15)))

print(iucn_levelplot1)
print(iucn_levelplot2)
# iucn_ln_pts <- as.data.frame(rasterToPoints(all_iucn_ln_map))
# spp_richness_iucn_plot <- ggplot(iucn_ln_pts, aes(x = x, y = y)) +
#   ggtheme_map + 
#   geom_raster(aes(fill = layer), alpha = .7) +
#   scale_fill_continuous(low = 'white', high = 'blue') +
#   borders('world', color='gray40', fill='gray45', size = .1) +  # create a layer of borders
#   scale_x_continuous(breaks = seq(-180, 180, by = 30), expand = c(0, 2)) +
#   scale_y_continuous(breaks = seq( -90,  90, by = 30), expand = c(0, 2)) +
#   labs(title = 'IUCN Dataset: log-transformed relative species richness', x = NULL, y = NULL, fill = NULL) 
# 
# spp_richness_iucn_plot
# ggsave(file.path(dir_fig, 'all_iucn_ln_map.png'), height = 4, width = 8)

@@

### 2015-12-21 ms_0_main.Rmd
trying this again... cleaning up data_explore plots and .csvs

``` {r load data, echo = FALSE, warning = FALSE, message = FALSE}
### use data_explore.Rmd to create this file, if it does not yet exist.

data_file <- 'data/spp_list_w_area_percent_noextinct_2015.csv'
spp_map_pairs <- read.csv(data_file, stringsAsFactors = FALSE) %>%
  filter(!is.na(popn_category)) %>% ### filter out unassessed species
  filter(!is.na(sm_perc))           ### filter out NA maps (usually IUCN maps with no polygons)

>>>>

``` {r load data, echo = FALSE, warning = FALSE, message = FALSE}
### use data_explore.Rmd to create this file, if it does not yet exist.  This
### filters out presence = Extinct from IUCN cells.

data_file <- 'data/spp_list_w_area_all.csv'
spp_map_pairs <- read.csv(data_file, stringsAsFactors = FALSE) %>%
  filter(!is.na(popn_category)) %>% ### filter out unassessed species
  filter(!is.na(sm_perc))           ### filter out NA maps (usually IUCN maps with no polygons)
